### 2019-11-13 paper/conceptual_framework.tex
Discuss the role of sex selection in spacing
Focus is on the trade-off that parents face when deciding whether to engage in
sex selection, especially relating to birth spacing.

Discuss how education is related to sex selection.

Remove notes from paper no longer needed.

This section provides a conceptual framework for understanding what determines spacing 
between births.
Since education is closely linked to the use of sex selection, I focus especially on the
the effects of education on spacing.
I first discuss generic spacing decisions frameworks, and then turn to how son preference
and sex selection fits into the decisions on spacing.

In the simplest possible dynamic model of fertility the spacing decisions is trivial.
Assume that parents derive benefits from having their children around, rather than from 
the births themselves, that spacing of children by itself does impact parental utility, 
and that the parents perfectly control when to have children. 
In this set up, the optimal decision is then to have the children as early as possible 
\citep{Newman1988}.
The simplest model clearly does not fit the observed spacing patterns, but the question 
is why.

@@

The expected effect of economies of scale in childrearing is to shorten the interval between 
births.
Childbearing and rearing is intensive in the mother's time, but taking care of two young
children is not twice as time consuming as taking care of one child 
\citep{Vijverberg1982,Espenshade1984}.
Hence, if childbearing and rearing requires the mother to reduce her participation in productive 
activities, parents can lower the cost of children by shorten birth spacing and, thereby, reduce
the amount of time the mother has to be away from productive activities.
In this set-up women with higher education will both have lower fertility and shorter
spacing than women with lower education because of the higher opportunity cost of time 
\citep{Ross1974,Newman1981}.
The extent to which this type of economies of scale is important depends on the extent to
which having children affects the mother's productive activities.
If she mainly works at home there likely is going to be less of an incentive to space 
children for this reason than if she is part of the labor force outside the home.
Similarly, higher household income, holding the mother's income contanst, reduces the incentive 
to space births for this reason.

In the absense of perfect capital markets, theory suggests that the steeper the parents' income 
profile over time, the longer the spacing between births 
\citep{Heckman1975,Wolpin1984,Newman1988}.
The idea is that although parents would prefer to have their children as early as possible to
enjoy them for longer there is also a direct cost to children.
At the early stage of life income is low, which means that the marginal utility of consumption 
is higher making parents less willing to give up consumption now to have a child, compare to 
later when they know that their income will be higher.
There is evidence that---at least for first births---there is, indeed, an effect of the income
profile over time \citep{Newman1984,Happel1984}.

@@

What are the effects of the introduction of sex selection?
The introduction of sex selection means that parents now have the option to avoid
birth of children with an undesired sex.
This impacts the observed birth spacing since, as discussed above, each sex-selective 
abortions increases the interval between births by six to twelve months.
The question is who is likely to use sex selection, when, and why?

[could I simply copy the results from my WB paper's theory section?]

There is clear evidence that in India the use of sex selection increases with education
\citep{das_gupta97,retherford03b,jha06,Guilmoto2009a,Bongaarts2013,Portner2015b,Jayachandran2017}.
There are two potential reasons for this.%
\footnote{
It is also possible that the effect is driven by a stronger son preference, although 
higher education women have shown declining stated son preference \citep{bhat03,pande07}.
}
First, women with more education generally have lower fertility either because of the
higher opportunity cost of time or because of other changes in the preferred number of
children.
Second, women with more education also tend to live in households with higher income,
which better enables them to access prenatal sex determination.
[It does not seem possible to cleanly separate out these differences]

should have something about how short spacing to conception could still be important.
Maybe even more so since health outcomes for short spacing birth likely has improved
over time.

[Notes/lit review from paper]

The sex composition of previous children affects both the timing of births and the use of 
sex-selective abortions 
\citep{retherford03b,jha06,Bhalotra2008,abrevaya09,Portner2015b,Kumar2016,Soest2018}.
I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
Finally, urban women have both lower fertility and, presumably, better access to
sex selection, both of which lead to higher use of sex-selective abortions than in 
rural areas \citep{retherford03b,jha06,Portner2015b}.

The longer average spacing and the relatively higher number of very long birth intervals 
seem to run counter to a general reduction in the length of women's reproductive spans 
found in earlier research \citep{Padmadas2004}.

Finally, we know less about what determines spacing behavior in developing countries than 
in developed countries, and with declining fertility and increasing numbers of women 
entering the labor force, understanding how couples make timing decisions is necessary 
for the design of suitable policies \citep{Portner2018}.%
\footnote{
It is clear that birth spacing does respond to policies in both developed and 
developing countries \citep{Pettersson-Lidbom2009,Todd2012,Meckel2015,Ghosh2018}.
}

>>>>

This section provides a conceptual framework for understanding what determines spacing 
between births.
Since sex selection is closely linked to education, I focus especially on the
the effects of education on spacing.
I first discuss a generic spacing decisions frameworks, and then turn to how son 
preference and sex selection fits into the decisions on spacing.

In the simplest possible dynamic model of fertility the spacing decision is trivial.
Assume that parents derive benefits from having their children around, rather than from 
the births themselves, that spacing of children by itself does impact parental utility, 
and that the parents perfectly control when to have children. 
In this set up, the optimal decision is then to have the children as early as possible 
\citep{Newman1988}.
The simplest model clearly does not fit the observed spacing patterns, but the question 
is why.

@@

The expected effect of economies of scale in childrearing is to shorten the interval between 
births.
Childbearing and rearing is intensive in the mother's time, but taking care of two young
children is not twice as time consuming as taking care of one child 
\citep{Vijverberg1982,Espenshade1984}.
Hence, if childbearing and rearing requires the mother to reduce her participation in productive 
activities, parents can lower the cost of children by shorten birth spacing and, thereby, reduce
the amount of time the mother has to be away from productive activities.
In this set-up women with higher education will both have lower fertility and shorter
spacing than women with lower education because of the higher opportunity cost of time 
\citep{Ross1974,Newman1981,Newman1984}.
The extent to which this type of economies of scale is important depends on the extent to
which having children affects the mother's productive activities.
If she mainly works at home there likely is going to be less of an incentive to space 
children for this reason than if she is part of the labor force outside the home.
Similarly, higher household income, holding the mother's income contanst, reduces the incentive 
to space births for this reason.

In the absense of perfect capital markets, theory suggests that the steeper the parents' income 
profile over time, the longer the spacing between births 
\citep{Heckman1976,Wolpin1984,Newman1988}.
The idea is that although parents would prefer to have their children as early as possible to
enjoy them for longer there is also a direct cost to children.
At the early stage of life income is low, which means that the marginal utility of consumption 
is higher making parents less willing to give up consumption now to have a child, compare to 
later when they know that their income will be higher.
There is evidence that---at least for first births---there is, indeed, an effect of the income
profile over time \citep{Newman1984,Happel1984}.

@@

With the introduction of sex selection parents can now avoid the birth of children with 
an undesired sex, although at the cost of the procedure and having to wait longer until 
the next child if the fetus is aborted.
The use of sex selection impacts the observed birth spacing since, as discussed 
above, each sex-selective abortions increases the interval between births by six 
to twelve months.
The question is who is likely to use sex selection, when, and why?

Theory suggests that the lower the desired fertility the more likely is the use of sex 
selection \citep{Portner2015b}.
This is driven by simple probabilities:
without use of sex selection, the probability of not having a son are 48.8 percent when
only having one child, 23.8 percent for two children, 11.6 percent for three children, 
5.7 percent for four children, 2.8 percent for five children, and 1.4 percent for six 
children.
Furthermore, given a desired level of fertility, the higher the parity, the higher the
likelihood of using sex selection.
Unless parents have a specific preference for a first-born boy, the costs associated with 
sex selection means that it is better to try for a boy without sex selection first and 
then use at later parities in the absence of a son as you get closer to the desired 
fertility.

The higher the cost of prenatal sex determination and the abortion, both in direct cost
and the possibility of criminal charges, the less use we should see.
Similarly, any potentially negative effects on the mother and offspring from repeated
abortions would also tend to reduce the use of sex selection---provided, obviously, that
parents are aware of these costs.
Finally, the more abortions between births, the less the opportunity there is to take
advantage of any economies of scale in childrearing.

The use of sex selection in India increases with education and is higher in urban
areas than in rural
\citep{das_gupta97,retherford03b,jha06,Guilmoto2009a,Bongaarts2013,Portner2015b,
Jayachandran2017}.
Women with more education generally have lower desired fertility, either because of the
higher opportunity cost of time or because of other changes in the preferred number of
children.%
\footnote{
It is also possible that the effect is driven by a stronger son preference, although 
higher education women have shown declining stated son preference \citep{bhat03,pande07}.
}
Women with more education also tend to live in households with higher income,
which better enables them to access prenatal sex determination and lowers the relative
costs of using sex selection (both direct and indirect, such the decline in 
economies of scale with repeated abortions).

Given that each abortion increases the spacing between births, it is possible that
parents may try to conceive even sooner than in the absence of sex selection.
By conceiving sooner, they would lower the expected cost of foregone economies of scale in 
childrearing.
This is especially relevant for better educated women, who, as discussed above, have
lower mortality risk for offspring born after short spacing.
If health outcomes for short spacing births has improved over time---as is likely---we
should see an even stronger push toward shorter spacing for women using sex selection.

[Need some kind of summary of spacing decisions and how they are likely to change
over time and with increased education]
